A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2018-11-27 01_Introduction.Rmd
more work on the introduction

This articles employs some of the methods that arose within archaeology's contemplation of *Cultural transmission theory* to trace the development of four major Bronze Age burial traditions in Europe. The cultural-historical evolution of archaeological cultures and ethnic groups during the European Bronze Age is complex and it's reconstruction an ongoing endeavour of archaeology. Within this history of constant emergence, displacement and demise many different forms of burial traditions were practised. Most of them can be distinguished by the criteria *burial construction* and *treatment of the body*. Bronze Age burials are either flat graves or part of a burial mound, and they contain either untreated or cremated human remains. Openly available bulk radiocarbon data now allows to reconstruct the trends within the distribution of these complex behaviours to test classical archaeological narratives about their respective prevalence. The comparison of macroregional trends also opens the perspective on long term patterns of interaction intensity on a continental scale.

In a nutshell, the classic narrative of the development of Bronze Age burial rites in Europe presented e.g. by @hardingEuropeanSocietiesBronze2000a (following @hauslerBestattungssittenFruhenBronzezeit1977; @hauslerGrabUndBestattungssitten1994; @hauslerTotenorientierungUndGeographischer1996)  can be summarised as follows: In the early and middle Bronze Age inhumations dominated in different variations. Cremation burials in this time only occur in the Hungarian lowlands. Burial mounds are concentrated in parts of the Balkans as well as in eastern, western and northern Europe, while in central and southern Europe flat graves - mostly burials - predominate. In the Middle Bronze Age, barrows gained importance in Western and Central Europe -- a phenomenon highly linked to the emergence of the Tumulus culture in Central Europe. In the late Bronze Age, the Urnfield period, cremation became the most common burial custom in the context of Urnfield culture.

>>>>

This articles employs some of the methods that arose within archaeology's contemplation of *Cultural transmission theory* to trace the development of four major Bronze Age burial traditions in Europe. The cultural-historical evolution of archaeological cultures and ethnic groups during the European Bronze Age is complex and it's reconstruction an ongoing endeavour of archaeology. Within this history of constant emergence, displacement and demise many different forms of burial traditions were practised. Most of them can be distinguished by the criteria *burial construction* and *treatment of the body*. Bronze Age burials are either flat graves or part of a burial mound, and they contain either untreated or cremated human remains. In a nutshell, the classic narrative of the development of these rites in Europe presented e.g. by @hardingEuropeanSocietiesBronze2000a (following @hauslerBestattungssittenFruhenBronzezeit1977; @hauslerGrabUndBestattungssitten1994; @hauslerTotenorientierungUndGeographischer1996) can be summarised as follows: In the early and middle Bronze Age inhumations dominated in different variations. Cremation burials in this time only occur in the Hungarian lowlands. Burial mounds are concentrated in parts of the Balkans as well as in eastern, western and northern Europe, while in central and southern Europe flat graves -- mostly burials -- predominate. In the Middle Bronze Age, barrows gained importance in Western and Central Europe. This is highly linked to the emergence of the *Tumulus culture* in Central Europe. In the late Bronze Age, the Urnfield period, cremation became the most common burial custom in the context of *Urnfield culture*.

Hardings narrative is, even when taking into account it's enormous simplification, partially misleading. Cremation, for example, was a lot more common already in the early and middle Bronze Age, though not in the context of the later Urnfield culture (e.g. in the Benelux area [@fokkensBronzeAgeLow2013] or Great-Britain [@robertsBritainIrelandBronze2013]). Beyond simple research progress, one additional reason why major misconceptions like this can appear and thrive might be a lack of unified data repositories and universally acknowledged proxies for the distribution of cultural phenomena in European archaeology: Instead of directly querrying quantitative data the typical process for the creation of large scale overviews is to excerpt the literature of all relevant local research traditions. The present article in contrast explores whether openly available bulk radiocarbon data could allow to reconstruct the trends within the distribution of burial rites. This could set the precedent for for a more reproducible and statistically sound reconstruction of the main shifts. The quantitative comparison of macroregional trends also opens the perspective on long term patterns of interaction intensity on a continental scale.
